### 2014-08-28 paper/paper.Rmd
filled in the gender section

The OECD describes the PISA survey as ``the world's global metric for quality, equity and efficiency in school education" (citep[p. 3]{PISA2014a}). The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across the 65 countries and economies. Students are examined on how well they can apply the knowledge they learned in school to applications outside of school. Additional information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affects their performance in school. Schools principals responded to a questionnaire covering the school system and learning experiences for students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

@@

### Gender gap in math is not universal but the reading gap is

@@

      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
  geom_point(aes(y=maxfemale), colour="pink") + 
  scale_y_continuous("High Math") + coord_flip() + theme_bw()

@@

      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
  geom_point(aes(y=minfemale), colour="pink") + 
  scale_y_continuous("Low Math") + coord_flip() + theme_bw()

@@

   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")

@@

student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
  scale_y_continuous("Science Score Gap") + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()

@@

```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}
student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("gendermathmap.pdf", width=12, height=7)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("genderreadmap.pdf", width=12, height=7)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
```

### School start age affects performance

Broken out by country we have computed the average math test score by age that the children started school. In additon, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, eg in Japan all children start at age 6.

@@

>>>>

The OECD describes the PISA survey as ``the world's global metric for quality, equity and efficiency in school education" (citep[p. 3]{PISA2014a}). The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across the 65 countries and economies. Students are examined on how well they can apply the knowledge they learned in school to applications outside of school. The scores reported range between 0-1000. Additional information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affects their performance in school. Schools principals responded to a questionnaire covering the school system and learning experiences for students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

@@

### The gender gap in math is not universal but the reading gap is, in favor of girls

For each country, average scores on math and reading are calculated separately for boys and girls. These averages are differenced to compute the "gap" between boys and girls. T-statistics were calculated to determine the statistical significance of the difference. The results are plotted in the graphs below. Each dot corresponds to a difference between girls and boys scores, for each country. Color indicates statistical significance (blue in favor of boys, pink in favor of girls and green means there was no significant difference). The size of the dot indicates the proportion of boys in the sample size. For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys. The maps color the countries by the presence of a gender gap.

The countries are sorted from biggest gap to smallest gap. Most countries have a gender gap in math scores in favor of boys. Colombia has the biggest gender gap, about 30 points on average boys do better than girls. It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.  Four countries have gender gaps in favor of girls, and even more surprising, these are middle eastern, or Muslim countries.

Conversely, though, when we look at reading scores the story is entirely different. Universally, girls score significantly better than boys. Again, the size of the gap is small, between 10 and 80 points.

Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys score. There is far more individual variability than the overall average differences.

@@

#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
#qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
#  xlab("") +
#  geom_point(aes(y=maxfemale), colour="pink") + 
#  scale_y_continuous("High Math") + coord_flip() + theme_bw()

@@

#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
#qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
#  xlab("") +
#  geom_point(aes(y=minfemale), colour="pink") + 
#  scale_y_continuous("Low Math") + coord_flip() + theme_bw()

@@

@@

#student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
#qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
#  xlab("") +
#  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
#  scale_y_continuous("Science Score Gap") + 
#  scale_size("Prop male") +
#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()

@@

```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("gendermathmap.pdf", width=12, height=7)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("genderreadmap.pdf", width=12, height=7)
#ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
#  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
#  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
#  new_theme_empty + theme(legend.position="none")
```

### Starting school at an older age generated lower average math score.

For each country we have computed the average math test score by age that the children started school. In additon, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, eg in Japan all children start at age 6.

@@

### Play with the data yourself

Shiny app

### Acknowledgements

Software packages
